---
title: "Germination fluidics"
author: "Jay T. Lennon"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

Distribution of events following virospore germination

## Clear workspace and set directory

```{r setup}
rm(list = ls())
getwd()
setwd("~/GitHub/viro.fluidics/")
```

## Load packages and functions

```{r}
require("png")
require("dplyr")
require("grid")
#require("plyr")
#require("tidyverse")
#require("bbmle")
```

# Load data
```{r}
fluid <- read.csv("data/fluidics.csv", header = TRUE, sep = ",", na.strings = "N/A")
head(fluid)
```

# Check data
```{r}
# Get unique categories from spore.fate
unique_categories <- unique(fluid$spore.fate)
print(unique_categories)
```

# Create new variable: time between germination and lysis = "death interval"
```{r}
fluid <- fluid %>%
  mutate(death.interval = lyse.time - germ.time)
```

# Remove "inconclusive" observation from spore.fate
```{r}
fluid <- fluid %>%
  filter(spore.fate != "Inconclusive")
```

# Proportions of spore.fate
```{r}
# Create count table
fate.count.table <- table(fluid$spore.fate, fluid$treat)

# View the table
fate.count.table 

# Create proportion table
fate.prop.table <- prop.table(spore.fate.table, margin = 2)

# View the table
fate.prop.table

# Reorder
fate.prop.table <- fate.prop.table[c("Divided", "Stayed", "Lysed"), ]
```

# Contingency test of spore.fate
```{r}
# Overall Fisher's exact test
fisher.fate <- fisher.test(spore.fate.table)
fisher.fate # p-value < 2.2e-16

# Pairwise Fisher's exact test

# Initialize vectors to store results
spore_fate_categories <- rownames(spore_fate_table)
pairwise_p_values <- c()

# Loop through each spore.fate category
for (category in spore_fate_categories) {
  # Create a 2x2 table: Current category vs. all other categories
  binary_table <- rbind(
    c(spore_fate_table[category, "control"], sum(spore_fate_table[, "control"]) - spore_fate_table[category, "control"]),
    c(spore_fate_table[category, "infect"], sum(spore_fate_table[, "infect"]) - spore_fate_table[category, "infect"])
  )
  
  # Perform Fisher's exact test
  test_result <- fisher.test(binary_table)
  
  # Store the p-value
  pairwise_p_values <- c(pairwise_p_values, test_result$p.value)
}

# Adjust p-values using Benjamini-Hochberg correction
adjusted_p_values <- p.adjust(pairwise_p_values, method = "BH")

# Combine results into a data frame
results <- data.frame(
  spore_fate = spore_fate_categories,
  p_value = pairwise_p_values,
  adjusted_p_value = adjusted_p_values
)

# View the results
results

#  spore_fate      p_value    adjusted_p_value
#   Divided    8.899003e-22     1.334850e-21
#   Lysed      1.057010e-31     3.171031e-31
#   Stayed     1.342470e-03     1.342470e-03
```

# Contingency bar plot
```{r}
categories <- c("Divided", "Stayed", "Lysed")
control.prop <- fate.prop.table[,1]
infect.prop <- fate.prop.table[,2]

# Combine data into a matrix for plotting
prop.data <- rbind(control.prop, infect.prop)

# Plot settings
barplot(
  prop.data,
  beside = TRUE,
  col = c("skyblue", "salmon"),
  names.arg = categories,
  legend.text = c("Unifected", "Infected"),
  #args.legend = list(x = "topright", bty = "n"),
  args.legend = list(x = "top", bty = "n"),
  ylim = c(0, 1),
  ylab = "Proportion",
  xlab = "Spore Fate",
  las = 1)  # Rotate y-axis tick labels)
  
abline(h = 0, col = "black", lwd = 1.5)
#  main = "Proportion of Spore Fate by Treatment"
```

# Distributions of continuous measures
```{r}
# Germination time (minutes)
#summary(fluid$germ.time)
#sd(fluid$germ.time, na.rm = TRUE)
hist(fluid$germ.time, 
     breaks = 20, # Adjust the number of bins
     main = "Histogram of germination time", 
     xlab = "Germination time (min)", 
     col = "lightgreen", 
     border = "darkgreen")


# Lysis time (minutes)
summary(fluid$lyse.time)
sd(fluid$lyse.time, na.rm = TRUE)
hist(fluid$lyse.time, 
     breaks = 20, # Adjust the number of bins
     main = "Histogram of lyse time", 
     xlab = "Lysis time (min)", 
     col = "pink", 
     border = "red")


# Cell length (um)
summary(fluid$lyse.time)
sd(fluid$cell.size, na.rm = TRUE)
hist(fluid$cell.size, 
     breaks = 20, # Adjust the number of bins
     main = "Histogram of cell size", 
     xlab = "Cell size (um)", 
     col = "lightblue", 
     border = "blue")

# Death interval 
summary(fluid$death.interval)
sd(fluid$death.interval, na.rm = TRUE)
hist(fluid$death.interval, 
     breaks = 20, # Adjust the number of bins
     main = "Histogram of death interval", 
     xlab = "Death interval (min)", 
     col = "orange", 
     border = "brown")


```

```{r}
# Subset data for control and infect
cont.germ <- fluid$germ.time[fluid$treat == "control"]
infect.germ <- fluid$germ.time[fluid$treat == "infect"]

# Define the range of the x-axis
x_range <- range(c(cont.germ, infect.germ), na.rm = TRUE)

hist(
  cont.germ,
  breaks = 20,  # Adjust the number of bins as needed
  col = rgb(0, 0, 1, 0.5),  # Semi-transparent blue
  xlim = x_range,
  ylim = c(0, 0.01),
  xlab = "Germination time (min)",
  ylab = "Density",
  main = "",
  freq = FALSE,
  border = "blue",
  las = 1
)

# Add the second histogram
hist(
  infect.germ,
  breaks = 20,  # Match the number of bins with the first histogram
  col = rgb(1, 0, 0, 0.25),  # Semi-transparent red
  add = TRUE,  # Overlay this histogram on the existing plot
  border = "red",
  freq = FALSE
)

# Add the density curve for the first vector
#lines(density(cont.germ, na.rm = TRUE, bw = 50), col = "blue", lwd = 2)
lines(density(cont.germ[cont.germ > 5], 
          na.rm = TRUE, bw = 50), col = "blue", lwd = 2)

# Add the density curve for the second vector
lines(density(infect.germ[infect.germ > 5], 
          na.rm = TRUE, bw = 50), col = "red", lwd = 2)


# Add a legend to distinguish the histograms
legend(
  x = "top",  # Positions the legend at the top middle
  legend = c("Uninfected", "Infected"),
  fill = c(rgb(0, 0, 1, 0.5), rgb(1, 0, 0, 0.25)),
  border = c("blue", "red"),
  bty = "n",  # Removes the box around the legend
  horiz = TRUE  # Makes the legend horizontal for better alignment
)


# Kolmogorov-Smirnov test
ks.germ <- ks.test(cont.germ, infect.germ)
print(ks.germ)
```



```{r}
# Load required libraries
if (!requireNamespace("evd", quietly = TRUE)) install.packages("evd")
if (!requireNamespace("bbmle", quietly = TRUE)) install.packages("bbmle")
library(evd)
library(bbmle)

# Input the dataset (replace this with your actual data or read from a file)
data <- read.table(header = TRUE, text = "
treat germ
infect 70
infect 26
infect 19
infect 68
infect 19
infect 194
infect 19
infect 195
infect 23
infect 48
infect 24
infect 24
infect 20
infect 12
infect 42
infect 6
infect 36
infect 61
infect 18
infect 21
infect 149
infect 88
infect 44
infect NA
infect 52
infect 60
infect 35
infect 271
infect 20
infect 31
infect 103
infect 88
infect 21
infect 24
infect 108
infect 22
infect 102
infect 16
infect 50
infect 105
infect 84
infect 176
infect 14
infect 39
infect 33
infect 23
infect 17
infect 27
infect 39
infect 31
infect 32
infect 20
infect 21
infect 153
infect 49
infect 49
infect 203
infect 33
infect 24
infect 64
infect 31
infect 28
infect 26
infect 120
infect 24
infect NA
infect 65
infect 30
infect 28
infect 25
infect 119
infect 24
infect NA
infect 23
infect 15
infect 19
infect 366
infect 35
infect 73
infect 67
infect 48
infect 57
infect 49
infect 52
infect 15
infect 21
infect 380
infect 107
infect 111
infect 24
infect 352
infect NA
infect 86
infect 31
infect 21
infect 201
infect 23
infect 32
control 84
control NA
control NA
control NA
control 165
control NA
control 30
control NA
control 183
control NA
control 4
control 44
control 30
control 19
control 12
control 6
control 76
control 4
control 90
control NA
control 4
control 165
control 188
control 30
control 32
control 128
control 39
control 36
control 126
control 19
control 159
control 98
control NA
control 48
control 103
control 4
control 20
control 43
control 109
control 102
control 7
control NA
control 140
control 64
control 36
control 116
control 79
control 20
control 39
control 32
control 139
control 28
control NA
control 53
control 9
control 25
control 54
control 10
control 41
control 66
control 188
control 38
control 118
control 52
control 55
control NA
control 101
control 35
control 12
control 173
control 211
control 5
control 25
control 231
control 149
control NA
control 217
control 134
control 50
control 189
control 70
control 117
")

# Remove rows with NA values
data <- na.omit(data)

gev_likelihood_treatment <- function(shape, location, scale, treat_shape = 0, treat_location = 0, treat_scale = 0) {
  # Apply treatment effects
  data$treat_effect <- ifelse(data$treat == "infect", 1, 0)
  shape_eff <- shape + treat_shape * data$treat_effect
  location_eff <- location + treat_location * data$treat_effect
  scale_eff <- scale + treat_scale * data$treat_effect
  
  # Validate parameters element-wise
  if (any(scale_eff <= 0)) {
    print("Invalid: Scale must be > 0")
    return(Inf)
  }
  
  max_germ <- max(data$germ, na.rm = TRUE)
  if (any(shape_eff < 0 & max_germ > (location_eff + scale_eff / abs(shape_eff)))) {
    print("Invalid: Tail constraint violated")
    return(Inf)
  }
  
  # Safely compute log-likelihood
  ll <- tryCatch({
    densities <- mapply(
      function(x, loc, sc, sh) {
        if (sc <= 0) return(Inf)  # Ensure valid scale
        dgev(x, shape = sh, loc = loc, scale = sc, log = TRUE)
      },
      x = data$germ,
      loc = location_eff,
      sc = scale_eff,
      sh = shape_eff
    )
    print(densities)  # Debugging
    sum(densities, na.rm = TRUE)
  }, error = function(e) {
    print("Error in dgev:")
    print(e)
    return(Inf)
  })
  
  if (!is.finite(ll)) {
    print("Likelihood is not finite")
    return(Inf)
  }
  
  -ll  # Return negative log-likelihood
}

gev_likelihood_treatment(
  shape = 0.05,
  location = 41.5,
  scale = 20,
  treat_shape = 0,
  treat_location = 0,
  treat_scale = 0
)

null_model <- mle2(
  gev_likelihood_treatment,
  start = list(shape = 0.05, location = 41.5, scale = 20),
  method = "L-BFGS-B",
  lower = list(shape = -0.1, location = 10, scale = 5),
  upper = list(shape = 0.1, location = 300, scale = 50),
  control = list(trace = 3)
)

summary(null_model)
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  # Calculate effective parameters
  shape_eff <- shape + treat_shape * data$treat_effect
  location_eff <- location + treat_location * data$treat_effect
  scale_eff <- scale + treat_scale * data$treat_effect
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  # Calculate effective parameters
  shape_eff <- shape + treat_shape * data$treat_effect
  location_eff <- location + treat_location * data$treat_effect
  scale_eff <- scale + treat_scale * data$treat_effect
  
    # Enforce scale > 0
  if (any(scale_eff <= 0)) return(Inf)
  
  # Enforce valid ranges for shape and data bounds
  max_germ <- max(data$germ, na.rm = TRUE)
  if (any(shape_eff < 0 & max_germ > (location_eff + scale_eff / abs(shape_eff)))) return(Inf)
  
  # Compute log-likelihood
  ll <- sum(dgev(data$germ, shape = shape_eff, loc = location_eff, scale = scale_eff, log = TRUE))
  -ll # Return negative log-likelihood for minimization
}

# Fit models with different treatment effects
models <- list(
  "Null (No Treatment Effect)" = mle2(
    gev_likelihood,
    start = list(shape = 0.1, location = 50, scale = 20),
    method = "L-BFGS-B",
    lower = list(shape = -1.5, location = 0, scale = 0.01),
    upper = list(shape = 1.5, location = max(data$germ, na.rm = TRUE), scale = 100)
  ),
  "Shape Affected by Treatment" = mle2(
    gev_likelihood,
    start = list(shape = 0.1, location = 50, scale = 20, treat_shape = 0),
    method = "L-BFGS-B",
    lower = list(shape = -1.5, location = 0, scale = 0.01, treat_shape = -1),
    upper = list(shape = 1.5, location = max(data$germ, na.rm = TRUE), scale = 100, treat_shape = 1)
  ),
  "Location Affected by Treatment" = mle2(
    gev_likelihood,
    start = list(shape = 0.1, location = 50, scale = 20, treat_location = 0),
    method = "L-BFGS-B",
    lower = list(shape = -1.5, location = 0, scale = 0.01, treat_location = -50),
    upper = list(shape = 1.5, location = max(data$germ, na.rm = TRUE), scale = 100, treat_location = 50)
  ),
  "Scale Affected by Treatment" = mle2(
    gev_likelihood,
    start = list(shape = 0.1, location = 50, scale = 20, treat_scale = 0),
    method = "L-BFGS-B",
    lower = list(shape = -1.5, location = 0, scale = 0.01, treat_scale = -50),
    upper = list(shape = 1.5, location = max(data$germ, na.rm = TRUE), scale = 100, treat_scale = 50)
  ),
  "All Parameters Affected by Treatment" = mle2(
    gev_likelihood,
    start = list(shape = 0.1, location = 50, scale = 20, treat_shape = 0, treat_location = 0, treat_scale = 0),
    method = "L-BFGS-B",
    lower = list(shape = -1.5, location = 0, scale = 0.01, treat_shape = -1, treat_location = -50, treat_scale = -50),
    upper = list(shape = 1.5, location = max(data$germ, na.rm = TRUE), scale = 100, treat_shape = 1, treat_location = 50, treat_scale = 50)
  )
)



# Compare models using AIC
aic_values <- AICtab(models, weights = TRUE)

# Display results
print(aic_values)


# Compare models using AIC
aic_values <- AICtab(models, weights = TRUE)

# Display results
print(aic_values)



```
